### 2018-01-26 manuscript.rmd
Edits throughout the text.

Catriona Munro^1^\*^,^**, Stefan Siebert^1,2^\*, Felipe Zapata^3^, Mark Howison^4,5^, Alejandro Damian Serrano^1,10^, Samuel H. Church^1,6^, Freya E. Goetz^1,7^, Philip R. Pugh^8^, Steven H.D. Haddock^9^, Casey W. Dunn^10^

@@

\*\* Corresponding author, catriona_munro@brown.edu

@@

Siphonophores are a diverse group of hydrozoans (Cnidaria) that are found at all depths of the ocean - from the surface, like the familiar Portuguese man of war, to the deep sea. Siphonophores play an important role in ocean ecosystems, and are among the most abundant gelatinous predators. A previous phylogenetic study based on two ribosomal RNA genes provided insight into the internal relationships between major siphonophore groups, however there was little support for many deep relationships within the clade Codonophora. Here, we present a new siphonophore phylogeny based on new transcriptome data from 30 siphonophore species analyzed in combination with 13 publicly available genomic and transcriptomic datasets. We use this new phylogeny to reconstruct several traits that are central to siphonophore biology, including sexual system (monoecy vs. dioecy), gain and loss of zooid types, life history traits, and habitat. The phylogenetic relationships in this study are largely consistent with the previous phylogeny, but we find strong support for new clades within Codonophora that were previously unresolved. These results have important implications for trait evolution within Siphonophora, including favoring the hypothesis that monoecy arose at least twice.

@@

Siphonophores remain poorly known, in large part because they are fragile and difficult to collect. They have, however, been of great interest for more than 150 years due to their unique structure and development [@mapstone2014global;@mackie1987siphonophore]. Like many other cnidarians, they are colonial: they grow by incomplete asexual reproduction. Each colony arises from a single embryo that forms the protozooid, the first body. One or two growth zones (Fig. 2) then arise that asexually produce other genetically identical zooids that remain attached [@carre1969etude; @carre1991complete; @carre1967; @carre1995ordre]. These zooids are each homologous to a solitary animal, but are physiologically integrated [@totton1965synopsis; @mackie1987siphonophore; @dunn2006evolution]. Siphonophores differ significantly from other colonial animals in terms of colony structure and development -- their zooids are highly functionally specialized and arranged in precise, repeating, species-specific patterns [@Beklemishev1969;@cartwright2010character]. The functions that zooids are specialized for include feeding, reproducing, or swimming (Fig. 2) [@dunn2006evolution].

![Photographs of living siphonophores. They correspond to the clades shown in Figure 3 as follows: Cystonectae (A-B), Calycophorae (C-G), Apolemiidae (H), and Clade A within Euphysonectae (I-K). (A) *Rhizophysa eysenhardtii*, scale bar = 1 cm. (B) *Bathyphysa conifera*, scale bar = 2cm. (C) *Hippopodius hippopus*, scale bar = 5 mm. (D) *Kephyes hiulcus*, scale bar = 2 mm. (E) *Desmophyes haematogaster*, scale bar = 5 mm. (F) *Sphaeronectes christiansonae*, scale bar = 2 mm. (G) *Praya dubia*, scale bar = 4 cm . (H) *Apolemia* sp., scale bar = 1 cm. (I) *Lychnagalma utricularia*, scale bar = 1 cm.  (J) *Nanomia* sp., scale bar = 1 cm. (K) *Physophora hydrostatica*, scale bar = 5 mm. Photo credits: S. Siebert (C,H,I,K), S. Haddock (A,D,E,F), R. Sherlock (B), MBARI (G), C. Munro (J)](figures/Figure1.jpg)

Understanding the unique ecology, morphology, and development of siphonophores requires a well-resolved phylogeny of the group. The relationship of siphonophores to other hydrozoans has been difficult to resolve [@cartwright2010character; @Cartwright2008kl; @zapata2015phylogenomic], but there has been progress on their internal relationships. A phylogeny [@Dunn:2005dy] based on two genes (16S, 18S) from 52 siphonophore taxa addressed several long standing questions about siphonophore biology. Including the relationships of the three historically recognised groups, Cystonectae, Physonectae, and Calycophorae. Cystonectae was found to be sister to all other siphonophores, while Calycophorae were nested within "Physonectae". The name Codonophora was given to this clade of "Physonectae" and Calycophorae [@Dunn:2005dy].

Major questions remained after this early work, though. There was, in particular, little support for important deep relationships within Codonophora. Understanding these relationships is key to resolving the evolution of several traits of importance, including sexual systems (monoecy versus dioecy) and the gain and loss of particular zooids, such as palpons (Fig. 2). Here we present a broadly sampled phylogenetic analysis of Siphonophora that considers transcriptomic data from 33 siphonophore species and 10 outgroup species (2 outgroups were subsequently excluded due to poor sampling). Using 1,071 genes, we find strong support for many relationships found in the earlier phylogeny [@Dunn:2005dy], and also provide new resolution for key relationships that were unresolved in that previous study. Using this phylogeny, we reconstruct the evolutionary history of characters central to the unique biology of siphonophores, including zooid type, life history traits, and habitat.

![Schematic of the siphonophore *Nanomia bijuga*, oriented with the anterior of the colony at the top of the page, and the ventral side to the left. Adapted from http://commons.wikimedia.org/wiki/File:Nanomia_bijuga_whole_animal_and_growth_zones.svg, drawn by Freya Goetz. (A) Overview of the whole mature colony. (B) Inset of the pneumatophore and nectosomal growth zone. A series of buds give rise to nectophores. (C) Inset of the siphosomal growth zone. Probuds subdivide to give rise to zooids in repeating-units (cormidia). The gastrozooid (specialized feeding polyp) is the posterior-most zooid within each cormidium.](figures/Figure2.pdf)

@@

All scripts for the analyses are available in a git repository at [https://github.com/caseywdunn/siphonophore_phylogeny_2017](https://github.com/caseywdunn/siphonophore_phylogeny_2017). The most recent commit at the time of the analysis presented here was `r system("git log | head -n 1", intern=TRUE) %>% str_replace("commit ", "")`.

@@

When possible, specimens were starved overnight in filtered seawater at temperatures close to ambient water temperatures at the time of specimen collection. mRNA was extracted directly from tissue using a variety of methods (Supplementary data 1): Magnetic mRNA Isolation Kit (NEB, #S1550S), Invitrogen Dynabeads mRNA Direct Kit (Ambion, #61011), Zymo Quick RNA MicroPrep (Zymo #R1050), or from total RNA after Trizol (Ambion, #15596026) extraction and through purification using Dynabeads mRNA Purification Kit (Ambion, #61006).  In case of anticipated very small total RNA quantities, only a single round of bead purification was performed. Extractions were performed according to the manufacturer's instruction. All samples were DNase treated (TURBO DNA-free, Invitrogen #AM1907; or on column DNase treatment with Zymo Quick RNA MicroPrep). Libraries were prepared for sequencing using the Illumina TruSeq RNA Sample Prep Kit (Illumina, #FC-122-1001, #FC-122-1002), the Illumina TruSeq Stranded Library Prep Kit (Illumina, #RS-122-2101) or the NEBNext RNA Sample Prep Master Mix Set (NEB, #E6110S). We collected long read paired end Illumina data for *de novo* transcriptome assembly. In the case of large tissue inputs, libraries were sequenced separately for each tissue, subsequently subsampled and pooled *in silico*. Libraries were sequenced on the HiSeq 2000, 2500, and 3000 sequencing platforms. Summary statistics for each library are given in the file Supplementary data 2. All sequence data have been deposited in the NCBI sequence read archive (SRA) with Bioproject accession number PRJNA255132.

@@

We used the Swofford-Olsen-Waddell-Hillis (SOWH) test [@swofford1996phylogenetic] to evaluate two hypotheses (Fig. 3C, S2): (i) "Physonectae" is monophyletic [@totton1965synopsis]; (ii) monoecious species are monophyletic [@Dunn:2005dy]. As the sexual system of *Rudjakovia* sp. is unclear, we carried out two tests of the monophyly of monoecy, one with *Rudjakovia* sp. included as a monoecious species, and one without (Fig. 3C, S2). We used SOWHAT [@doi:10.1093/sysbio/syv055] dev. version 0.39 (commit fd68ef57) to carry out the SOWH tests in parallel with the default options and an initial sample size of 100 (analysis code can be found in the git repository). For each hypothesis we defined a topology with a single constrained node that was inconsistent with the most likely topology (Fig. 3). We used a threshold for significance of 0.05 and following the initial 100 samples, we evaluated the confidence interval around the p-value to determine if more samples were necessary.

@@

Most relationships received strong support across analysis methods (Fig. 3A), with a couple of localized exceptions (Fig. 3B). The phylogenetic relationships recovered in this study are largely consistent with those found in a previous study based on two genes (16S and 18S ribosomal RNA) [@Dunn:2005dy]. Relationships that receive strong support in both studies include the placement of Cystonectae as sister to Codonophora (the clade that includes all other siphonophores), the placement of Apolemiidae as sister to all other codonophorans, and the placement of Calycophorae within the paraphyletic "Physonectae". Multiple nodes that were not resolved in the previous two-gene analysis receive strong support in the present 1,071-gene transcriptome analyses. There is strong support for Pyrostephidae as sister to all other non-apolemiid codonophorans. Within the clade that is sister to Pyrostephidae, we find two main clades, Calycophorae and a clade we here name Euphysonectae (Fig. 3A). It includes the remaining non-apolemiid, non-pyrostephid "Physonectae". We define Euphysonectae as the clade consisting of *Agalma elegans* and all taxa that are more closely related to it than to *Diphyes dispar*.

@@

Within Clade A, *Physophora gilmeri* along with *Lychnagalma utricularia* (Fig. 1I) (both not included in the previous phylogeny) are sister to Agalmatidae, a clade restricted to *Agalma*, *Athorybia*, *Melophysa*, *Halistemma* and *Nanomia* [@Dunn:2005dy; @pugh2006taxonomic]. In the rDNA study, *P. hydrostatica* (the presumed sister species to *P. gilmeri*) was sister to Forskaliidae with low support. The position of *Cordagalma cordiforme* (= *Cordagalma ordinatum*) [@pugh2016synopsis] was previously unresolved, while in this analysis *Cordagalma* sp. is in a clade with *Forskalia asymmetrica*, falling outside of Agalmatidae. Placement of *Cordagalma* outside Agalmatidae is consistent with previous analyses of morphological and molecular data [@pugh2006taxonomic; @Dunn:2005dy].

Within Calycophorae, taxon sampling is shallower here than in the previous study. The calycophoran relationships that can be investigated, however, are in broad agreement with the previous analysis. Calycophorans have in the past been split into two groups, prayomorphs and diphyomorphs, based on morphology after @mackie1987siphonophore. As in the previous study, the results presented here indicate that the prayomorphs are paraphyletic with respect to the diphyomorphs. *Craseoa lathetica* and *Desmophyes* sp. are sister to *Hippopodius hippopus* in this study, while in the previous study, the relationship between *C. lathetica* and the clade including *H. hippopus* was unresolved.

@@

Using the Swofford-Olsen-Waddell-Hillis (SOWH) test [@swofford1996phylogenetic], we evaluated the following three alternative phylogenetic hypotheses against the most likely tree topology (Fig. 3C): (i) monophyletic Physonectae, (ii) and (iii) monophyletic monoecious siphonophores, with and without *Rudjakovia* sp. respectively. In all three tests the alternative hypothesis was rejected (p-value <0.01, confidence interval: <0.001 - 0.03, Fig. S2).

The broad sampling approach of this phylogeny provides new evidence for the relationships between major siphonophore clades within Codonophora, specifically between Pyrostephidae, Calycophorae, and the newly named Euphysonectae. This opens up new questions about key relationships within both Calycophorae and Euphysonectae -- where future transcriptome sampling efforts should be focused. Within Euphysonectae, two clades (Clade A and Clade B) are hypothesized, however there is weaker support for Clade B (Fig. 3A, 3B). Expanding sampling of species that probably fall in Clade B, including other *Erenna* species, rhodaliids, and relatives of Undescribed sp L, will greatly expand our understanding of these two groups and perhaps provide evidence of Clade B synapomorphies. Similarly, within Calycophorae, increased taxon sampling is needed. This study, and the previous phylogenetic study [@Dunn:2005dy], suggest that the prayomorphs are paraphyletic, but for slightly different reasons given the different sampling of the analyses. In @Dunn:2005dy, a clade of prayomorphs including *Praya dubia* (Fig. 1G), *Nectadamas diomedeae*, and *Nectopyramis natans* (not included in this study) were found to be sister to all other calycophorans, while in this study, the prayomorph *Lilyopsis fluoracantha* (not included in the previous study) is found in a clade including diphyomorph calycophorans that is sister to all other prayomorphs. Expanded taxon sampling, particularly *P. dubia* or a nectopyramid, but also extensive sampling across the major prayomorph and diphyomorph groups, will expand our understanding of relationships within Calycophorae. This will be especially important for understanding trait evolution within Calycophorae,  for example, the release of eudoxids (Fig. 4), or the arrangement of male and female zooids along the stem (see section 3.2 below).

@@

In all siphonophores, each gonophore (sexual medusa that produces gametes) is either male or female. Within each siphonophore species, colonies are either monoecious (male and female gonophores are on the same colony) or dioecious (male and female gonophores are on different colonies). Previous analyses suggested that the common ancestor of siphonophores was dioecious, and was consistent with a single gain of monoecy within Codonophora and no secondary losses [@Dunn:2005dy]. The better-resolved tree in the current analyses indicates that the evolution of monoecy is more complicated than this. The two clades of monoecious siphonophores, Calycophorae and Clade A (Fig. 3A), do not form a monophyletic group. This is because Clade B, which contains dioecious species, is also descended from their most recent common ancestor. SOWH tests strongly reject the placement of the monecious clades Calycophorae and Clade A as a group that excludes Clade B (Figs. 3C and S2). The positions of the only two taxa from Clade B in the previous analysis [@Dunn:2005dy], *Erenna* and *Stephalia*, were unresolved in that previous study. This difference in conclusions regarding trait evolution, therefore, does not reflect a contradiction between alternative strongly supported results, but the resolution of earlier polytomies in a way that indicates there has been homoplasy in the evolution of monoecy.

The distribution of monoecy is consistent with two scenarios (Fig. 4). In the first, there is a single shift from dioecy to monoecy along the branch that give rise to the most recent common ancestor of Calycophorae and Euphysonectae, followed by a shift back to dioecy along the branch that gave rise to Clade B. In the second, monoecy arose twice -- once along the branch that gave rise to Clade A and again along the branch that gave rise to Calycophorae.

Ancestral character state reconstructions favor the hypothesis that monoecy arose twice (Fig. 5A), once in Calycophorae and once in Clade A. This is consistent with differences in the arrangements of male and female gonophores in the two clades. In Clade A, male and female zooids are found within the same cormidium (a single reiterated sequence of zooids along the stem, see Fig. 2). In these species, the male and female zooids are placed at very different but well defined locations within the cormidium. Meanwhile in calycophorans, each cormidium bears either male or female gonophores. In this form of monoecy, the male and female cormidia can either be in an alternating pattern, or there can be several male or female cormidia in a row. In either case, male and female zooids are found at the same corresponding locations within the cormidia. In sum, homoplasy in sexual system evolution along with variation in the spatial arrangement of gonophores within a colony suggest that siphonophores have evolved different ways to be monoecious. The sexual system and cormidial arrangement of *Rudjakovia* is undescribed, although preliminary observations suggest that this species may be monoecious and that monoecy arose a third time in the Pyrostephidae. A detailed redescription of *Rudjakovia* would help clarify this.

@@

Within Calycophorae there are some variations in sexual mode -- in *Sulculeolaria* (not included in this phylogeny) colonies appear to consist of only one sex, however they are monoecious and protandrous, with female gonophores developing after the release of male gonophores [@carre1979]. Environmental influences may also play a role in determining the expressed sex. Colonies of the calycophoran *Chelophyes appendiculata* collected in the field always bear both male and female gonophores whereas when kept in culture only gonophores of one sex are maintained [@Carre:2000hb]. This suggests a high plasticity of the sexual state in some calycophoran taxa and underlines the need for caution when evaluating the state of this character in rarely collected species.

@@

One of the most striking aspects of siphonophore biology is their diversity of unique zooid types [@Beklemishev1969; @cartwright2010character]. For example, members of *Forskalia* have 5 basic zooid types (nectophore, gastrozooid, palpon, bract, and gonophore), and in some species, a total of 9 when considering zooid subtypes (4 types of bract, male & female gonophores)[@pugh2003revision]. Here we reconstruct the evolutionary origins of several zooid types on the present transcriptome tree (Fig. 4).

Nectophores (Fig. 2) are non-reproductive propulsive medusae. In Codonophora, the nectophores are localized to a region known as the nectosome (Fig. 2B), which has its own growth zone, and are used for coordinated colony-level swimming. Planktonic cystonects like *Bathyphysa sibogae* and *Rhizophysa filiformis* (Fig. 1A) instead move through the water column using repeated contraction and relaxation of the stem, and use modified flattened gastrozooids with wings (called ptera) to increase surface area and prevent colony sinking [@biggs1976siphonophore].  Nectophores are also present within the gonodendra (reproductive structures) of cystonects, and are thought to propel the gonodendra when they detach from the colony [@totton1965synopsis;@Totton1960]. It is not clear whether the nectophores found within the siphosome of the cystonects are homologous to the nectophores borne on the nectosome of codonophorans. Similarly, the homology of the special nectophore associated with gonophores of the calycophoran *Stephanophyes superba* is also unclear [@chun1891canarischen]. As such, we only consider the evolution of the nectosome in this study, and not the presence/absence of nectophores. The present analyses, as well as the analyses of @Dunn:2005dy, are consistent with a single origin of the nectosome (Fig. S5).

>>>>

Catriona Munro^1^\*^,^‡, Stefan Siebert^1,2^\*, Felipe Zapata^3^, Mark Howison^4,5^, Alejandro Damian Serrano^1,10^, Samuel H. Church^1,6^, Freya E. Goetz^1,7^, Philip R. Pugh^8^, Steven H.D. Haddock^9^, Casey W. Dunn^10^

@@

\‡ Corresponding author, catriona_munro@brown.edu

@@

Siphonophores are a diverse group of hydrozoans (Cnidaria) that are found at all depths of the ocean - from the surface, like the familiar Portuguese man of war, to the deep sea. Siphonophores play important roles in ocean ecosystems, and are among the most abundant gelatinous predators. A previous phylogenetic study based on two ribosomal RNA genes provided insight into the internal relationships between major siphonophore groups, however there was little support for many deep relationships within the clade Codonophora. Here, we present a new siphonophore phylogeny based on new transcriptome data from 30 siphonophore species analyzed in combination with 13 publicly available genomic and transcriptomic datasets. We use this new phylogeny to reconstruct several traits that are central to siphonophore biology, including sexual system (monoecy vs. dioecy), gain and loss of zooid types, life history traits, and habitat. The phylogenetic relationships in this study are largely consistent with the previous phylogeny, but we find strong support for new clades within Codonophora that were previously unresolved. These results have important implications for trait evolution within Siphonophora, including favoring the hypothesis that monoecy arose at least twice.

@@

Siphonophores remain poorly known, in large part because they are fragile and difficult to collect. They have, however, been of great interest for more than 150 years due to their unique structure and development [@mapstone2014global;@mackie1987siphonophore]. Like many other cnidarians, they are colonial: they grow by incomplete asexual reproduction. Each colony arises from a single embryo that forms the protozooid, the first body. One or two growth zones (Fig. 2) then arise and asexually produce other genetically identical zooids that remain attached [@carre1969etude; @carre1991complete; @carre1967; @carre1995ordre]. These zooids are each homologous to a solitary animal, but are physiologically integrated [@totton1965synopsis; @mackie1987siphonophore; @dunn2006evolution]. Siphonophores differ significantly from other colonial animals in their colony structure and development -- their zooids are highly functionally specialized and arranged in precise, repeating, species-specific patterns [@Beklemishev1969;@cartwright2010character]. Zooids are specialized for a range of functions, including feeding, reproduction, or swimming (Fig. 2) [@dunn2006evolution].

![Photographs of living siphonophores. Colored circles correspond to the clades shown in Figure 3 as follows: Cystonectae (A-B), Calycophorae (C-G), Apolemiidae (H), and Clade A within Euphysonectae (I-K). (A) *Rhizophysa eysenhardtii*, scale bar = 1 cm. (B) *Bathyphysa conifera*, scale bar = 2cm. (C) *Hippopodius hippopus*, scale bar = 5 mm. (D) *Kephyes hiulcus*, scale bar = 2 mm. (E) *Desmophyes haematogaster*, scale bar = 5 mm. (F) *Sphaeronectes christiansonae*, scale bar = 2 mm. (G) *Praya dubia*, scale bar = 4 cm . (H) *Apolemia* sp., scale bar = 1 cm. (I) *Lychnagalma utricularia*, scale bar = 1 cm.  (J) *Nanomia* sp., scale bar = 1 cm. (K) *Physophora hydrostatica*, scale bar = 5 mm. Photo credits: S. Siebert (C,H,I,K), S. Haddock (A,D,E,F), R. Sherlock (B), MBARI (G), C. Munro (J)](figures/Figure1.jpg)

Understanding the unique ecology, morphology, and development of siphonophores requires a well-resolved phylogeny of the group. The relationship of siphonophores to other hydrozoans has been difficult to determine [@cartwright2010character; @Cartwright2008kl; @zapata2015phylogenomic], but there has been progress on their internal relationships. A phylogeny [@Dunn:2005dy] based on two genes (16S, 18S) from 52 siphonophore taxa addressed several long standing questions about siphonophore biology, including the relationships of the three historically recognised groups, Cystonectae, Physonectae, and Calycophorae. Cystonectae was found to be sister to all other siphonophores, while Calycophorae were nested within "Physonectae". The name Codonophora was given to this clade of "Physonectae" and Calycophorae [@Dunn:2005dy].

Major questions remained after this early work, though. In particular, there was little support for important deep relationships within Codonophora. Understanding these relationships is key to resolving the evolution of several traits of importance, including sexual systems (monoecy versus dioecy) and the gain and loss of particular zooids, such as palpons (Fig. 2). Here we present a broadly sampled phylogenetic analysis of Siphonophora that considers transcriptomic data from 33 siphonophore species and 10 outgroup species (2 outgroups were subsequently excluded due to poor sampling). Using 1,071 genes, we find strong support for many relationships found in the earlier phylogeny [@Dunn:2005dy], and also provide new resolution for key relationships that were unresolved in that previous study. Using this phylogeny, we reconstruct the evolutionary history of characters central to the unique biology of siphonophores, including zooid type, life history traits, and habitat.

![Schematic of the siphonophore *Nanomia bijuga*, oriented with the anterior of the colony at the top of the page, and the ventral side to the left. Adapted from http://commons.wikimedia.org/wiki/File:Nanomia_bijuga_whole_animal_and_growth_zones.svg, drawn by Freya Goetz. (A) Overview of the whole mature colony. (B) Inset of the pneumatophore and nectosomal growth zone. A series of buds give rise to nectophores. (C) Inset of the siphosomal growth zone. Probuds subdivide to give rise to zooids in repeating units (cormidia). The gastrozooid (specialized feeding polyp) is the posterior-most zooid within each cormidium.](figures/Figure2.pdf)

@@

All scripts for the analyses are available in a git repository at [https://github.com/caseywdunn/siphonophore_phylogeny_2017](https://github.com/caseywdunn/siphonophore_phylogeny_2017). The most recent commit at the time of the analysis presented here was `r system("git log | head -n 1", intern=TRUE) %>% str_replace("commit ", "")` with tag XXXXXXXXXX.

@@

When possible, specimens were starved overnight in filtered seawater at temperatures close to ambient water temperatures at the time of specimen collection. Subsequently, mRNA was extracted directly from tissue using a variety of methods (Supplementary data 1): Magnetic mRNA Isolation Kit (NEB, #S1550S), Invitrogen Dynabeads mRNA Direct Kit (Ambion, #61011), Zymo Quick RNA MicroPrep (Zymo #R1050), or from total RNA after Trizol (Ambion, #15596026) extraction and through purification using Dynabeads mRNA Purification Kit (Ambion, #61006).  In case of anticipated very small total RNA quantities, only a single round of bead purification was performed. Extractions were performed according to the manufacturer's instruction. All samples were DNase treated (TURBO DNA-free, Invitrogen #AM1907; or on column DNase treatment with Zymo Quick RNA MicroPrep). Libraries were prepared for sequencing using the Illumina TruSeq RNA Sample Prep Kit (Illumina, #FC-122-1001, #FC-122-1002), the Illumina TruSeq Stranded Library Prep Kit (Illumina, #RS-122-2101) or the NEBNext RNA Sample Prep Master Mix Set (NEB, #E6110S). We collected long read paired end Illumina data for *de novo* transcriptome assembly. In the case of large tissue inputs, libraries were sequenced separately for each tissue, subsequently subsampled and pooled *in silico*. Libraries were sequenced on the HiSeq 2000, 2500, and 3000 sequencing platforms. Summary statistics for each library are given in the file Supplementary data 2. All sequence data have been deposited in the NCBI sequence read archive (SRA) with Bioproject accession number PRJNA255132.

@@

We used the Swofford-Olsen-Waddell-Hillis (SOWH) test [@swofford1996phylogenetic] to evaluate two hypotheses (Fig. 3C, S2): (i) "Physonectae" is monophyletic [@totton1965synopsis]; (ii) monoecious species are monophyletic [@Dunn:2005dy]. As the sexual system of *Rudjakovia* sp. is unclear, we carried out two tests of the monophyly of monoecy, one with *Rudjakovia* sp. included as a monoecious species, and one with it considered dioecious (Fig. 3C, S2). We used SOWHAT [@doi:10.1093/sysbio/syv055] dev. version 0.39 (commit fd68ef57) to carry out the SOWH tests in parallel, using the default options and an initial sample size of 100 (analysis code can be found in the git repository). For each hypothesis we defined a topology with a single constrained node that was inconsistent with the most likely topology (Fig. 3). We used a threshold for significance of 0.05 and following the initial 100 samples, we evaluated the confidence interval around the p-value to determine if more samples were necessary.

@@

The phylogenetic relationships recovered in this study received strong support across analysis methods (Fig. 3A), with a couple of localized exceptions (Fig. 3B). Most clades are consistent with those found in a previous study based on two genes (16S and 18S ribosomal RNA) [@Dunn:2005dy]. Relationships that receive strong support in both include the placement of Cystonectae as sister to Codonophora (the clade that includes all other siphonophores), the placement of Apolemiidae as sister to all other codonophorans, and the placement of Calycophorae within the paraphyletic "Physonectae". Multiple nodes that were not resolved in the previous two-gene analysis receive strong support in the present 1,071-gene transcriptome analyses. There is strong support for Pyrostephidae as sister to all other non-apolemiid codonophorans. Within the clade that is sister to Pyrostephidae, we find two main clades, Calycophorae and a clade we here name Euphysonectae (Fig. 3A). It includes the remaining non-apolemiid, non-pyrostephid "Physonectae". We define Euphysonectae as the clade consisting of *Agalma elegans* and all taxa that are more closely related to it than to *Diphyes dispar*.

@@

Within Clade A, *Physophora gilmeri* along with *Lychnagalma utricularia* (Fig. 1I) (both not included in the previous phylogeny) are sister to Agalmatidae, a clade restricted to *Agalma*, *Athorybia*, *Melophysa*, *Halistemma* and *Nanomia* [@Dunn:2005dy; @pugh2006taxonomic]. In the rDNA study, *P. hydrostatica* (the presumed sister species to *P. gilmeri*) was sister to Forskaliidae with low support. The position of *Cordagalma cordiforme* (= *Cordagalma ordinatum*) [@pugh2016synopsis] was previously unresolved, while in this analysis *Cordagalma* sp. is in a clade with *Forskalia asymmetrica*, falling outside of Agalmatidae. Placement of *Cordagalma* outside Agalmatidae is consistent with previous analyses of molecular and morphological data [@pugh2006taxonomic; @Dunn:2005dy].

Within Calycophorae, taxon sampling is less comprehensive here than in the previous study. The calycophoran relationships that can be investigated, however, are in broad agreement with the previous analysis. Calycophorans have in the past been split into two groups, prayomorphs and diphyomorphs, based on morphology after @mackie1987siphonophore. As in the previous study, the results presented here indicate that the prayomorphs are paraphyletic with respect to the diphyomorphs. *Craseoa lathetica* and *Desmophyes* sp. are sister to *Hippopodius hippopus* in this study, while in the previous study, the relationship between *C. lathetica* and the clade including *H. hippopus* was unresolved.

@@

Using the Swofford-Olsen-Waddell-Hillis (SOWH) test [@swofford1996phylogenetic], we evaluated the following three alternative phylogenetic hypotheses against the most likely tree topology (Fig. 3C): (i) monophyletic Physonectae, (ii) monophyletic monoecious siphonophores including *Rudjakovia* sp. and (iii) monophyletic monoecious with dioecious *Rudjakovia* sp. <!--  XXX This needs to be fixed to coincide with the two hypotheses listed in section 2.3 XXX -->. In all three tests the alternative hypothesis was rejected (p-value <0.01, confidence interval: <0.001 - 0.03, Fig. S2).

The broad taxon sampling and more extensive gene sampling of this phylogeny provide new evidence for the relationships between major siphonophore clades within Codonophora, specifically between Pyrostephidae, Calycophorae, and the newly named Euphysonectae. This opens up new questions about key relationships within both Calycophorae and Euphysonectae -- where future transcriptome sampling efforts should be focused. Within Euphysonectae, two clades (Clade A and Clade B) are hypothesized, although there is weaker support for Clade B (Fig. 3A, 3B). Expanding sampling of species that probably fall in Clade B, including other *Erenna* species, rhodaliids, and relatives of Undescribed sp L, will greatly expand our understanding of these two groups and perhaps provide evidence of Clade B synapomorphies. Similarly, within Calycophorae, increased taxon sampling is needed. This study, and the previous phylogenetic study [@Dunn:2005dy], suggest that the prayomorphs are paraphyletic, but for slightly different reasons given the different sampling of the analyses. In @Dunn:2005dy, a clade of prayomorphs including *Praya dubia* (Fig. 1G), *Nectadamas diomedeae*, and *Nectopyramis natans* (not included in this study) were found to be sister to all other calycophorans, while in this study, the prayomorph *Lilyopsis fluoracantha* (not included in the previous study) is found in a clade including diphyomorph calycophorans that is sister to all other prayomorphs. Expanded transcriptome sequencing, particularly *P. dubia* or a nectopyramid, but also extensive sampling across the major prayomorph and diphyomorph groups, will expand our understanding of relationships within Calycophorae. This will be especially important for understanding trait evolution within Calycophorae,  for example, the release of eudoxids (Fig. 4), or the arrangement of male and female zooids along the stem (see section 3.2 below).

@@

In all siphonophores, each gonophore (sexual medusa that produces gametes) is either male or female. Within each siphonophore species, colonies are either monoecious (male and female gonophores are on the same colony) or dioecious (male and female gonophores are on different colonies). Previous analysis suggested that the common ancestor of siphonophores was dioecious, and was consistent with a single gain of monoecy within Codonophora and no secondary losses [@Dunn:2005dy]. The better-resolved tree in the current analyses indicates that the evolution of monoecy is more complicated than this. The two clades of monoecious siphonophores, Calycophorae and Clade A (Fig. 3A), do not form a monophyletic group. This is because Clade B, which contains dioecious species, is also descended from their most recent common ancestor. SOWH tests strongly reject the placement of the monecious clades Calycophorae and Clade A as a group that excludes Clade B (Figs. 3C and S2). The positions of the only two taxa from Clade B that were included in the previous analysis [@Dunn:2005dy], *Erenna* and *Stephalia*, were unresolved in that study. This difference in conclusions regarding trait evolution, therefore, does not reflect a contradiction between alternative strongly supported results, but the resolution of earlier polytomies in a way that indicates there has been homoplasy in the evolution of monoecy.

The distribution of monoecy is consistent with two potential scenarios (Fig. 4). In the first, there is a single shift from dioecy to monoecy along the branch that gave rise to the most recent common ancestor of Calycophorae and Euphysonectae, followed by a shift back to dioecy along the branch that gave rise to Clade B. In the second, monoecy arose twice: once along the branch that gave rise to Clade A and again along the branch that gave rise to Calycophorae.

Ancestral character state reconstructions favor the hypothesis that monoecy arose twice (Fig. 5A), once in Calycophorae and once in Clade A. This is consistent with differences in the arrangements of male and female gonophores in the two clades. In Clade A, male and female zooids are found within the same cormidium (a single reiterated sequence of zooids along the stem, see Fig. 2). In these species, the male and female zooids are placed at different but well defined locations within the cormidium. Meanwhile in calycophorans, each cormidium bears either male or female gonophores. In this form of monoecy, the male and female cormidia can either occur in an alternating pattern, or there can be several male or female cormidia in a row. In either case, male and female zooids are found at the same corresponding locations within the cormidia. In sum, homoplasy in sexual-system evolution along with variation in the spatial arrangement of gonophores within a colony suggest that siphonophores have evolved different ways to be monoecious. The sexual system and cormidial arrangement of *Rudjakovia* is undescribed, although preliminary observations suggest that this species may be monoecious and that monoecy arose a third time in the Pyrostephidae. A detailed redescription of *Rudjakovia* would help clarify this.

@@

Within Calycophorae there are additional variations of the sexual mode: in *Sulculeolaria* (not included in this phylogeny) colonies appear to consist of only one sex at a time. However they are monoecious and protandrous, with female gonophores developing after the release of male gonophores [@carre1979]. Environmental influences may also play a role in determining the expressed sex. Colonies of the calycophoran *Chelophyes appendiculata* collected in the field always bear both male and female gonophores, whereas when kept in culture only gonophores of one sex are maintained [@Carre:2000hb]. This suggests a high plasticity of the sexual state in some calycophoran taxa and underlines the need for caution when evaluating the state of this character in rarely collected species.

@@

One of the most striking aspects of siphonophore biology is their diversity of unique zooid types [@Beklemishev1969; @cartwright2010character]. For example, *Forskalia* and other physonects have at least 5 basic zooid types (nectophore, gastrozooid, palpon, bract, and gonophore), and in some species, there can be nine zooid subtypes (4 types of bract, male & female gonophores)[@pugh2003revision]. Here we reconstruct the evolutionary origins of several zooid types on the present transcriptome-based tree (Fig. 4).

Nectophores (Fig. 2) are non-reproductive propulsive medusae. In Codonophora, the nectophores are localized to a region known as the nectosome (Fig. 2B), which has its own growth zone, and they are used for coordinated colony-level swimming. Planktonic cystonects like *Bathyphysa sibogae* and *Rhizophysa filiformis* (Fig. 1A) instead move through the water column using repeated contraction and relaxation of the stem, and use a gas-filled float and modified flattened gastrozooids with wings (called ptera) to increase surface area and prevent colony sinking [@biggs1976siphonophore].  Nectophores are also present within the gonodendra (reproductive structures) of cystonects, and are thought to propel the gonodendra when they detach from the colony [@totton1965synopsis;@Totton1960]. It is not clear whether the nectophores found within the siphosome of the cystonects are homologous to the nectophores borne on the nectosome of codonophorans. Similarly, the homology of the special nectophore associated with gonophores of the calycophoran *Stephanophyes superba* is also unclear [@chun1891canarischen]. In this study, we only consider the evolution of the nectosome, and not the presence/absence of nectophores. The present analyses, as well as the analyses of @Dunn:2005dy, are consistent with a single origin of the nectosome (Fig. S5).

![Siphonophore phylogeny showing the distribution of the main anatomical characters and the bathymetric distributions of the different species. Bottom: siphonophore phylogeny, colored by clade. Middle panel: diagram showing the presence/absence of traits across Siphonophora, with the physical location of the trait shown on a schematic of *Nanomia bijuga* (schematic by Freya Goetz). Top: Bathymetric distribution of siphonophore species. *Physalia* illustration by Noah Schlottman, taken from http://phylopic.org/](figures/Figure4.pdf)
